---
title: "chapter6.Rmd"
author: "Matilda K"
date: "2023-12-11"
output: html_document
---
# Assignment 6
## 1. Meet and Repeat: PART I with RATS data
### 1.1. Reading data in
```{r read RATS in}
# Read the RATS data
RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", header = TRUE, sep = '\t')

# Look at the (column) names of BPRS
names(RATS)

# Look at the structure of BPRS
str(RATS)

# Print out summaries of the variables
summary(RATS)
```

### 1.2 Graphical displays of longitudinal data: The magical pivot_longer()
```{r graphical displays of long RATS}
# Access the packages dplyr and tidyr
library(dplyr)
library(tidyr)

# Factor variables ID and Group
RATS$ID <- factor(RATS$ID)
RATS$Group <- factor(RATS$Group)

# Convert to long form
RATSL <- pivot_longer(RATS, cols = -c(ID, Group), 
                      names_to = "WD",
                      values_to = "Weight") %>% 
         mutate(Time = as.integer(substr(WD,3,4))) %>%
         arrange(Time)

# Take a glimpse at the RATSL data
glimpse(RATSL)
```

### 1.3 Individuals on the plot
```{r plotting RATS}
#Access the package ggplot2
library(ggplot2)

# Draw the plot
ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") #+ 
  scale_y_continuous()
```
From this plot we can see that group 2 and 3 differ clearly from the group one already with starting weights. Group 1 also has more individuals in it. In group 2 there is an individual that seems to be an outlier with much higher weight throughout the study.

### 1.4 The Golden Standardise
```{r standardising and plotting}
library(dplyr)
library(tidyr)
# Standardise the variable RATSL
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdrats = (Weight - mean(Weight))/sd(Weight)) %>%
  ungroup()

# Glimpse the data
glimpse(RATSL)

# Plot again with the standardised bprs
library(ggplot2)
ggplot(RATSL, aes(x = Time, y = stdrats, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "standardized weight")
```
In the plot with standardized values the lines are more straight as the weight have been standardised in relation to all other weights. Whereas in the previous non standardised plot the weight had an increasing trend, in this plot some of the standardised weights have even a decreasing trend.

### 1.5 Summary graphs
```{r summary graphs of RATS}
library(dplyr)
library(tidyr)
# Summary data with mean and standard error of bprs by treatment and week 
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean = Weight, se = Weight ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSS)

# Plot the mean profiles
library(ggplot2)
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.8)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
```
In the mean profiles plot it is again visible that the group 1 is different from the groups 2 and 3 with lower mean weights.

### 1.6 Find the outlier
```{r find outlier}
library(dplyr)
library(tidyr)
# Create a summary data by Group and ID with mean as the summary variable (ignoring baseline week 0)
RATSL8S <- RATSL %>%
  #filter(week > 0) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSL8S)

# Draw a boxplot of the mean versus treatment
library(ggplot2)
ggplot(RATSL8S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), weeks 1-9")

# Create a new data by filtering the outlier and adjust the ggplot code the draw the plot again with the new data
RATSL8S1 <- filter(RATSL8S, mean <= 550)

# Glimpse the data
glimpse(RATSL8S1)
ggplot(RATSL8S1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), weeks 1-9")
```
In the first boxplot of the mean versus the group, the outlier in group 2 is visible. Looks like groups 1 and 3 also have outliers although less so that group 2. After removing the outlier of group 2 the box representing the deviation got much smaller.


### 1.7 Anova
```{r and Anova}
library(dplyr)
library(tidyr)
# Add the baseline from the original data as a new variable to the summary data
RATSL8S2 <- RATSL8S %>%
  mutate(baseline = RATS$WD1)

# Fit the linear model with the mean as the response 
fit <- lm(mean ~ baseline + Group, data = RATSL8S2)

# Compute the analysis of variance table for the fitted model with anova()
anova(fit)


```
Based on the anova results the baseline weights are strongly related to the mean weights in the following weeks. 


## 2. Meet and Repeat: PART II with BPRS data
### 2.1 Reading data in
```{r read BPRS in}
BPRS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt", sep =" ", header = T)

# Look at the (column) names of BPRS
names(BPRS)

# Look at the structure of BPRS
str(BPRS)

# Print out summaries of the variables
summary(BPRS)
```

### 2.2 Plots and Linear Mixed Effects Models
```{r bprsl plots}
library(dplyr)
library(tidyr)
# Factor treatment & subject
BPRS$treatment <- factor(BPRS$treatment)
BPRS$subject <- factor(BPRS$subject)

# Convert to long form
BPRSL <-  pivot_longer(BPRS, cols = -c(treatment, subject),
                       names_to = "weeks", values_to = "bprs") %>%
          arrange(weeks) #order by weeks variable

# Extract the week number
BPRSL <-  BPRSL %>% 
            mutate(week = as.integer(substr(weeks,5,5)))

# Take a glimpse at the BPRSL data
glimpse(BPRSL)

# Plot the data
library(ggplot2)

ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line()

ggplot(BPRSL, aes(x = week, y = bprs, group = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 60, 10)) +
  scale_y_continuous(name = "bprs") +
  theme(legend.position = "top")
  
  ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))

```

```{r regression model}
# create a regression model RATS_reg
BPRS_reg <- lm(bprs ~ week + treatment, data = BPRSL)

# print out a summary of the model
summary(BPRS_reg)
```
Based on the coefficient estimates of the summary statistics of the fitted model, higher number of week is associated with a lower likelihood for bprs being high. This coefficient is statistivally significant with the p-value of <2e-16. Therefore from the summary of the model one could say that the later the evaluation was done the lower the likelihood of them having high result in the BPRS. At this point it is good to keep in mind that this model assumes independence of the repeated measures which in this case is not very likely.

### 2.3 More appropriate models: random intercept and random slope model
```{r random intercept model}
library(lme4)

# Create a random intercept model
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)

# Print the summary of the model
summary(BPRS_ref)
```

```{r random slope model}
# create a random intercept and random slope model
library(lme4)
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)

# print a summary of the model
summary(BPRS_ref1)

# perform an ANOVA test on the two models
anova(BPRS_ref1, BPRS_ref)
```

Anova results show that the model with random intercept and slope (BPRS_ref1) with the p-value of 0.02636 is significantly better than the model with just random intercept (BPRS_ref).

### 2.4 Random Intercept and Random Slope Model with interaction

```{r Random Intercept and Random Slope Model with interaction}
# create a random intercept and random slope model with the interaction
library(lme4)
BPRS_ref2 <- lmer(bprs ~ week + treatment + (week | treatment), data = BPRSL, REML = FALSE)

# print a summary of the model
summary(BPRS_ref2)

# perform an ANOVA test on the two models
anova(BPRS_ref2, BPRS_ref1)

# draw the plot of BPRSL with the observed Weight values
ggplot(BPRSL, aes(x = week, y = bprs, group = subject)) +
  geom_line() +
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 60, 20)) +
  scale_y_continuous(name = "BPRS") +
  theme(legend.position = "top")
  

# Create a vector of the fitted values
Fitted <- fitted(BPRS_ref2)

library(dplyr)
library(tidyr)
# Create a new column fitted to RATSL
BPRSL <- BPRSL %>% mutate(Fitted = Fitted)

# draw the plot of BPRSL with the Fitted values of weight
library(ggplot2)
ggplot(BPRSL, aes(x = week, y = Fitted, group = subject)) +
  geom_line() +
  scale_x_continuous(name = "Time (weeks)") +
  scale_y_continuous(name = "BPRS") +
  theme(legend.position = "top")




```
I can't get this last plot to work correctly. Based on the anova test on the models BPRS_ref2 and BPRS_ref1 the BPRS_ref2 model did not improve from the first one so it can be discarded.